Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-24 manuscript/main.qmd
final proof reading & corrections

Here, we present a novel Bayesian hierarchical model [e.g. @Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled palaeoclimatic data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological palaeoclimatic proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients, that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitonti2011], lack. We test the robustness of this method using down-sampled, simulated latitudinal temperature gradients.

@@

Geochemical climatic proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Consequently, samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained (Fig. 1). For a detailed description of each proxy see @Hollis2019.

@@

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is assumed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.

@@

**K.** Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is broadly negative, i.e. temperature decreases with absolute latitude. This is achieved by setting $K \ge A$. The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 10, \alpha_{TN} = A, \beta_{TN}= \infty)$${#eq-K}
The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).

**M.** The steepness of the gradient is presumed to be highest in mid-latitudes; this is expressed with a normal prior on $M$ with the mean set to 42, i.e. $M$ of the modern SST gradient, and a moderately wide standard deviation of 10: $$M \sim N(42,10)$${#eq-M}

**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to limit very abrupt SST changes across latitudes on a global scale. A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$. We informed the prior distributions on $M$ and $B$ based on a provisional model run with the modern SST data.

@@

We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes, i.e. lower latitudes are sampled more frequently. For the largest sample size (n = 34), we used the latitudes of the EECO data set of this study in all iterations. For each latitude, we inferred a location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which is the average standard deviation of the EECO geochemical proxy data. With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, *ect*. We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is actual, average magnitude of uncertainty that the proxy compilation is afflicted with.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every $10^{th}$ iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients were modelled in one chain with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every $25^{th}$ iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the simulated gradients with an Eocene sampling distribution, a single chain with 250,000 iterations was used, thinned to 10,000 iterations after burn-in. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every $100^{th}$ iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Fig. S1).

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
Modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median and 95% CI of temperatures were then taken from all temperature estimates obtained at the latitudes of interest.

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20, and 34 (latitudes of EECO samples). Each row depicts a different latitudinal temperature gradient that represents idealised or observed climatic states: idealsed extreme icehouse, greenhouse, and extreme greenhouse gradients, and the modern gradient, which represents an icehouse state. The black line illustrates the original gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples (first three columns), and a single run with the EECO latitudinal sampling distribution (fourth column). To generate the random samples, noise from a normal distribution with a standard deviation of 3.8 was added to all temperatures. The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

Present-day latitudinal temperature gradient. The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise ($R^2 = `r round(R2_modern,2)`, N = 42,896)$. Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient. The darkness of the background shading indicates the point density.  
:::

Our Bayesian model is able to model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig. 2). Random latitudinal sampling results in accurate reconstructions for most random samples at sample sizes of 10 and 20 for the icehouse scenarios (median of $R^2$ > 0.9). Greenhouse scenarios perform somewhat worse due to the increased uncertainty at high latitudes (median of $R^2$ > 0.7 at sample sizes 10 and 20). A sampling distribution resembling that of the early Eocene dataset used in this study allows for accurate reconstruction of all scenarios, although the $R^2$ is still relatively low in the extreme greenhouse scenario, as a perfectly flat gradient prediction by the model would result in an $R^2$ of 0, despite the original gradient being very flat. This also explains the low lower bounds of the 95% credible intervals in the greenhouse scenarios.

@@

Estimates of the median, latitudinal sea surface temperature gradients of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model. The purple ribbon (shading) depicts the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible. Points within the plot depict the geochemical (e.g. TEX~86~) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted at the mean temperature values of their respective normal distributions.
:::

@@

Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimatic reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], and leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimatic reconstructions across the Phanerozoic. Our modelling framework offers a straightforward, efficient way of integrating ecological palaeoclimatic data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate. The model could easily be extended to include uncertainties on individual geochemical proxy data, or to variably weight proxy records classified as more or less reliable.

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but notably shallower when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, palaeoclimatic simulations from General Circulation Models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

@@

Temporal changes within the EECO [@Westerhold2018], and seasonality [@Keating-Bitonti2011; @Ivany2022] may also contribute to the large variability of the EECO proxy data. Based on the occurrence of heterotrophic carbonates, @Davies2019 suggest that mid- and high-latitude geochemical proxy data from the EECO may be biased towards summer temperatures. Some of the geochemical mid-latitude geochemical proxy data from @Hollis2019 may therefore suggest higher than actual mean annual temperatures, and the variability of temperature estimates from individual localities is higher in mid - high latitudes (Fig. S6). It is difficult to attribute this variability to seasonality alone, as temporal climate variability is also expected to be higher in mid and high latitudes[@Schwartz2008]. Critically, however, the mangrove data strongly supports our inference of a flattened gradient independent of the geochemical proxy record.

>>>>

Here, we present a novel Bayesian hierarchical model [e.g. @Gelman2013; @Mcelreath2018] that combines quantitative proxies and ecological constraints into a fully quantitative model of the latitudinal gradient of sea surface temperatures, bridging spatial gaps in sparsely sampled palaeoclimatic data. The Bayesian approach offers a powerful framework for integrating various sources of uncertainty and modelling complex hierarchical relationships, and is increasingly used in palaeoclimatic reconstructions [e.g. @Weitzel2019; @Yang2022; @Burgener2023]. This model expands upon existing, spatially explicit palaeoclimatic reconstructions  by allowing for the integration of (1) prior information based on physical principles and the observed modern sea surface temperature distribution, and of (2) geochemical and ecological palaeoclimatic proxies in a common, quantitative framework. We chose a generalised logistic function to accurately infer the shape of the temperature gradient despite a patchy latitudinal coverage. This choice is motivated by the flexibility and ability of this function to approximate a variety of nonlinear patterns in the underlying temperature gradients that other parametric approaches, such as lower order polynomials [e.g. @Bijl2009; @Keating-Bitonti2011], lack. We test the robustness of this method using down-sampled, simulated latitudinal temperature gradients.

@@

Geochemical climatic proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Consequently, samples labelled as "thermocline", or "sub-thermocline", were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained (Fig. 1). For a detailed description of each proxy see @Hollis2019.

@@

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures between 15.6&deg;C (lower temperature limit of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is assumed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.

@@

**K.** Input of solar energy decreases from the tropics to the poles. Hence, the latitudinal temperature gradient is broadly negative, i.e. temperature decreases with absolute latitude. This is achieved by setting $K \ge A$. The prior on the upper asymptote $K$ is a truncated normal distribution with the mean set to $K$ of the modern SST gradient, with a broad standard deviation: $$K \sim TN(\mu_{TN} = 28, \sigma_{TN} = 15, \alpha_{TN} = A, \beta_{TN}= \infty)$${#eq-K}
The distribution is truncated to the left at $\alpha_{TN}=A$, but not truncated to the right ($\beta_{TN}$).

**M.** A uniform prior is placed on the latitude of greatest steepness of the gradient, allowing it to be steepest anywhere between latitudes 0&deg; and 90&deg; absolute latitude, as this parameter may vary greatly depending on the climate state: $$M \sim Uniform(0,90)$${#eq-M}

**B.** The steepness or growth rate $B$ of the gradient is constrained to be $\ge0$ and to not be exceedingly high, as oceanic and atmospheric heat transfer is bound to limit very abrupt SST changes across latitudes on a global scale. A gamma-distributed prior of the form $$B \sim Gamma(\alpha_G=4.3,\beta_G=30)$${#eq-B}
was placed on $B$. The shape and rate parameters $\alpha_G$ and $\beta_G$ were chosen such that the highest prior density is at $B$ of the modern SST gradient, $0.11$. We informed the prior distribution on $B$ based on a provisional model run with the modern SST data.

@@

We created test data from these gradients as follows: We randomly sampled (1000 iterations) latitudes at sample sizes of 5, 10 and 20, with the probability of a latitude being sampled scaling with the decreasing surface area towards higher latitudes, i.e. lower latitudes are sampled more frequently. For the largest sample size (n = 34), we used the latitudes of the EECO data set of this study in all iterations. For each latitude, we took the location mean temperature from the gradients, adding random noise from a normal distribution with a standard deviation of 3.8, which corresponds to the average uncertainty associated with the EECO geochemical proxy data [@Hollis2019] . With that, we aim to simulate randomly distributed errors in the proxy data, which could arise from miscalibrations, measurement errors, seasonal effects, *ect*. We acknowledge that this approach cannot quantify the potential impact of systematic offsets that may bias all proxy data in the same direction, nor do we know whether a standard deviation of 3.8 is the actual average magnitude of uncertainty that the proxy compilation is afflicted with.

@@

We estimated the posterior distributions of the model parameters using a Markov chain Monte Carlo (MCMC) algorithm, written in R. Specifically, we sampled the unknown parameters $A$, $K$, $M$ and $B$ with Metropolis-Hastings, and used Gibbs sampling to estimate all other unknown parameters [see @Gilks1995; @Gelman2013]. Posterior inference on the modern gradient is based on four chains with 60,000 iterations each, 10,000 of which were discarded as burn-in. Every $10^{th}$ iteration was retained, resulting in a total of 20,000 iterations with low autocorrelation. The re-sampled, simulated gradients and the re-sampled, modern gradient were modelled in one chain  with 10,000 iterations for each of the 1,000 random samples. 5,000 iterations each were discarded as burn-in, and every $25^{th}$ iteration was kept, resulting in a total of 200,000 iterations across all 1,000 model runs. For the Eocene model, we ran four chains with 600,000 iterations each, discarding 100,000 as burn-in and keeping every $100^{th}$ iteration, as the hierarchical model structure results in higher autocorrelation of the chains. The Eocene posterior inference is thus based on a total of 20,000 iterations with low autocorrelation (effective multivariate sample size for $A$, $K$, $M$ and $B$ is \> 18,000). Trace plots of the MCMC chains indicate convergence and good mixing of the chains (Fig. S1).

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
Modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior samples. The median and 95% CI of temperatures were then taken from all temperature estimates obtained at the latitudes of interest.

@@

Model reconstructions of simulated latitudinal temperature gradients at various sample sizes. Each column depicts a different reconstruction for given sample sizes: 5, 10, 20 (randomly sampled latitudes), and 34 (latitudes of EECO samples). Each row depicts a different latitudinal temperature gradient that represents idealised or observed climatic states: idealised extreme icehouse, greenhouse, and extreme greenhouse gradients, and the modern gradient, which represents an icehouse state. The black line illustrates the original gradient. The blue line depicts the reconstructed gradient represented by the median sea surface temperature value estimated from 1,000 model runs with different random samples. To generate the random samples, different random noise from a normal distribution with a standard deviation of 3.8&degC; was added to each temperature. The blue shadings depict the 90%, 95%, and 99% credible intervals. Bold black text within each panel depicts the coefficient of determination ($R^2$) for estimating goodness of fit between the simulated and modelled gradient. The median (50%) $R^2$ value along with the 95% credible intervals from all model runs are shown. Each gradient is depicted in absolute latitude.
:::

@@

Present-day latitudinal temperature gradient. The present-day empirical latitudinal temperature gradient (median sea surface temperature) is depicted as a black line, and the gradient estimated by the Bayesian model is shown in turquoise ($R^2 = `r round(R2_modern,2)`, N = 42,896)$. Grey points depict the individual cell values of the Bio-ORACLE grid of mean sea surface temperatures, which were used to infer the empirical and the modelled gradient. Higher opacity of points indicates higher density of data (multiple overlapping points).
:::

Our Bayesian model is able to model a range of idealised temperature gradients, ranging from extreme icehouse to 'super greenhouse' scenarios (Fig. 2). Random latitudinal sampling results in accurate reconstructions for most random samples at sample sizes of 10 and 20 for the icehouse scenarios (median of $R^2$ > 0.9). Greenhouse scenarios perform somewhat worse due to the increased uncertainty at high latitudes (median of $R^2$ > 0.7 at sample sizes 10 and 20). A sampling distribution resembling that of the early Eocene dataset used in this study allows for accurate reconstruction of all scenarios, although the $R^2$ is still relatively low in the extreme greenhouse scenario, as a perfectly flat gradient, predicted by the model, would result in an $R^2$ of 0, despite the original gradient being very flat. This also explains the low lower bounds of the 95% credible intervals in the greenhouse scenarios.

@@

Estimates of the median, latitudinal sea surface temperature gradient of the early Eocene climatic optimum (purple line) and of the present-day (turquoise), both estimated with the Bayesian model. The purple ribbon (shading) depicts the 95% credible interval of the Eocene gradient, the uncertainty of the modern gradient is too low to be visible. Points within the plot depict the geochemical (e.g. TEX~86~) and ecological (e.g. mangroves) data. Geochemical data are plotted by their point estimate temperature value. Ecological data are plotted at the mean temperature values of their respective normal distributions.
:::

@@

Our results further exemplify how incorporating quantified ecological temperature constraints can provide more precise temperature reconstructions than geochemical proxies alone, adding to the advances in palaeoclimatic reconstructions achieved by integrating lithological data [@Scotese2021; @Burgener2023]. Combining the occurrences of climate-sensitive plant communities [@Greenwood1995], reptiles [@Markwick2007], and leaf shapes [@Peppe2011], with geochemical proxies offers substantial potential for improving quantitative palaeoclimatic reconstructions across the Phanerozoic. Our modelling framework offers a straightforward, efficient way of integrating ecological palaeoclimatic data with other proxy data: The hierarchical model structure accounts for variation of temperature estimates from proxies at individual localities, which is treated equivalent to the uncertainty associated with the ecological temperature proxies. A local temperature estimate, based on multiple geochemical proxies, thus has the same weight as a local temperature estimate obtained from the occurrence of a climate-sensitive plant community, whilst preserving the uncertainty associated with each estimate. The model could easily be extended to include uncertainties on individual geochemical proxy data (see Fig. S5), or to variably weight proxy records classified as more or less reliable.

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7--21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but notably shallower when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`&deg;C, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg;. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, palaeoclimatic simulations from General Circulation Models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

@@

Temporal changes within the EECO [@Westerhold2018], and seasonality [@Keating-Bitonti2011; @Ivany2022] may also contribute to the large variability of the EECO proxy data. Based on the occurrence of heterotrophic carbonates, @Davies2019 suggested that mid- and high-latitude geochemical proxy data from the EECO may be biased towards summer temperatures. Some of the geochemical mid-latitude geochemical proxy data from @Hollis2019 may therefore suggest higher than actual mean annual temperatures, and the variability of temperature estimates from individual localities is higher in mid - high latitudes (Fig. S6). It is difficult to attribute this variability to seasonality alone, as temporal climate variability is also expected to be higher in mid and high latitudes[@Schwartz2008]. Critically, however, the mangrove data strongly supports our inference of a flattened gradient independent of the geochemical proxy record.
